The CMS Data Aggregation System

### 2010-05-04 doc/paper.tex
More updates
git-svn-id: svn+ssh://svn.cern.ch/reps/CMSDMWM/DAS/trunk@558 4525493e-7705-40b1-a816-d608a930855b

The DAS core communicates with different data-services and retrieve
data on demand. This was done by invoking appropriate data-service APIs
upon user requests. All user queries were written to Analytics DB
for further analysis. The data-service output where re-mapped into
DAS notations and if necessary aggregated inside of DAS cache.
Since we didn't know a-priory what kind of data users will ask, 
what kind of queries they will place against the system and what
level of aggregation will be applied the choice
of RDMS as DAS back-end was problematic\footnote{In addition DAS 
didn't impose any requirements for transactions and persistency of the data.} 
and we look around for other options. Through the analysis of available IT solutions, 
such as file-based and memory caches, 
key-value databases, documented-oriented databases we made our choice in favor 
of the last technology. Among them we evaluated CouchDB \cite{CouchDB} and 
MongoDB \cite{MongoDB}. Both of them provide schema free document
storage, replication and fail-over features, but our choice in favor of 
MongoDB was obvious because its support of dynamic queries, 
full indexies, including on inner objects and embedded arrays,
and auto-sharding. Our prelimiary benchmarks shown that it can sustain
the desired load and size for our meta-data information. We use MongoDB for
the three main components: DAS cache, Mapping and Analytics databases. All of them
will be discussed in forthcoming section.

DAS Mapping DB holds mapping between data-service APIs and DAS keys. 
Upon identification of APIs which will participate in DAS activity 
we store to Mapping DB API name, input parameters, their type and accept 
values as regular expression patterns as well as mapping between data-service 
notations into DAS keys. This allows to map user input query parameters into
data-service API metrics as well as map back data-service output into DAS names.
For instance, the DBS system used $logical\_file\_name$ notation, while in PhEDEx
\cite{PhEDEx}
the same entity was named as $lfn$. Since both of them were referred to
logical file name, we map both notations into DAS key $file$.\footnote{We
also deal with physical file names, but still used $file$ to refer to them in
DAS notation, since simple regular expressions were used to identify $file$
meaning from the provided value.}
For each service we provide a mapping from its own data-format (JSON, XML, etc.) into
common data-format used in DAS.\footnote{We used JSON as a DAS data format.}
We also store a mapping between DAS keys and data-service API names to
establish appropriate calls to data-service upon provided DAS query. For example,
when user specifies $file$ in their DAS query, we invoke DBS listFiles and PhEDEx
fileReplicas APIs.

The DAS Analytics DB plays special role for DAS. It collects information
about user requests placed to the system. Each request was recorded. Upon its
decomposition into set of selection keys and conditions we also record
API name and API input parameters and keep updating their counter for repeated
calls. This allows to keep track of frequency of API calls and establish pre-fetch
strategy for most requested APIs.

The DAS caching system was used to store data-service results and aggregated
information in form of DAS records. As it shown in figure \ref{DAS_cache} 
the DAS cache server has dual purpose.
On one side it talks to DAS Analytics DB to trigger jobs for worker robots
who update cache with most popular queries, on another side it talks to DAS
cache back-end to either stored mapped results or merge them with 
existing DAS objects in a cache.

@@

Apart from previous DBS-QL syntax we decided to fully support free text-based queries for end-users.
All of them were transformed into MongoDB query syntax, who was used internally in
DAS core layer. It represents a set of selection keys and conditions as a JSON document.
For example, a query
$dataset\,\,site=T1\_CH\_CERN$
where transformed into 
$\{"fields" : ["dataset"], "spec" : \{ "site" :"T1\_CH\_CERN"\}\}$.
Such syntax was perfect fit for us, since it allows us to store all queries
as JSON documents into Analytics DB for further analysis. It is worthwhile to note that
MongoDB query syntax is very reach. It supports advanced queries with 
boolean expressions, conditional operators, regular expression, etc. providing
a real power for end-user queries.

\subsection{Data-services and aggregation}
DAS provided a pluggable system to add new CMS data-services. To do that we
identify appropriate APIs, input and output parameters, data-service
format and their mapping to DAS keys as well as relation to other sub-systems. 
This information was stored into DAS Mapping DB. A data-service plugin consists
of its DAS map and individual data parser. The DAS Mapping DB becomes an authorative
source of information about all data-services participating in DAS and their
relationships. Based on this information we were able to retrieve appropriate data
from different data-services, re-map them into DAS notations and
aggregate them on demand.
For instance, the Data Bookkeeping System (DBS) and data location system (PhEDEx)
provided information about CMS data files. In former case, DBS holds information about
files and their relationship to other physics objects, such as run and luminosity, 
while PhEDEx stores information about the file location, such as sites which
holds copy of the files and details about storage elements.
Upon a user request to find a file, we were able to identify appropriate set of
DBS and PhEDEx APIs, look-up data, re-format their output to DAS notations and
store results into DAS cache.
Once data were retreived from data-services an additional step to aggregate
matched objects in DAS cache were applied. For instance, both data-service outputs 
contained a file name in them. We use it as a common key to merge records in DAS cache.
This merged (aggregated) results represented a DAS data objects. Each DAS data
object contains a standard DAS header, which identify look-up time,
request url, input parameters and API method as well as response version, expiration
time and checksum of the output.

We used REST model \cite{REST} for all data requests to DAS.
It allows us to share core library and other components during development of 
various DAS clients, e.g. CLI, web interface, robots, etc. and hides
complexity of access to individual data-services and security policies.
For instance, to access on-line Run Summary information we were forced to use
Grid certificates. Instead of propagating user certificate to data-service
the DAS works as a proxy and use host certficiate to serve such requests.

\section{Results\label{Results}}
DAS was deployed plugins to work with the following CMS data-services:
\begin{itemize}
\item DBS \cite{DBS}, Data Bookkeeping System which collect information
about CMS meta-data;
\item PhEDEx \cite{PhEDEx}, Physics Experiment Data Export project which
provides the data placement and the file transfer system for the CMS experiment;
\item SiteDB \cite{SiteDB} is a catalogue of all CMS computing Tiers. 
It records the CMS resources at the site, the resources pledged for the 
future, and it keeps track of CMS personnel at each site, including the 
roles and duties they fulfill in the collaboration;
\item RunSummary DB \cite{RunSummary} collects information about run and triggers
conditions during data-taking;
\item LumiDB \cite{LumiDB} collects information about Luminosity conditions;
\item DQ \cite{DQ}, Data Quality data-service which collects information
about detector conditions during data taking;
\item Overview \cite{Overview}, collects information about CMS
transfer rate, etc.;
\item Dashboard \cite{Dashboard} project for LHC experiments aims to 
provide a single entry point to the monitoring data collected from the 
distributed computing systems of the LHC virtual organization.
\end{itemize}
Each data-service has its own scope, size and lifetime. For instance the data
in PhEDEx are transient due to constant migration of CMS data with daily DB size of
around few tens of GBs. The DBS system were divided into dozen of individual DBs, 
who has been used by different production teams for various bookkeeping tasks. Typical
size of individual DBS instance was around of fwe GBs, with overall DBS size of
around of 50 GBs.\footnote{This include table and index sizes.}

At the moment we don't have exact numbers for amount of data we will store to DAS,
but based on test performed during CRAFT09 exercise we produced 0.3PB of data which
translate into $\sim50$ GB of meta-data stored into DBS system. Based on this
we can anticipate to collect $\sim500$GB of meta-data per year during data-taking period
at LHC.
%In CRAFT09 acquisition era, Tier-0 produced a total of 6.03
%billion events, 330TB in 32 datasets:
%AlCa/Calibration RAW: 926 million events, 29TB
%Physics RAW: 1.24 billion events, 147TB
%PromptReco: 1.24 billion events, 111TB
%Bulk ALCARECO: 2.34 billion events, 11TB
%Express FEVT: 105 million events, 24TB
%Express ALCARECO: 153 million events, 0.99TB
%HLTMON FEVTHLTALL: 31 million events, 7.4TB
%Above event numbers have signicant overlap by design, 
%2.2 billion unique events, including 524 million cosmic muon
%triggers
% 0.3PB - 50GB meta-data in DBS
% 3PB/y - x GB

@@

We measured elapsed time as the following:
\begin{itemize}
\item[]
Elapsed time = retrieval time + parsing time + remapping time 
        + cache insertion/indexing time 
        + (aggregation time) + (output creation time)
\end{itemize}
Here the retrieval time is a time required to access data from remote data-services,
parsing time is a time required to read and parse received data, remapping time
is a time to convert notations used by data-service to DAS ones for every object
we parse, cache insertion and indexing time represents time spend to add objects into
the DAS cache, aggregation time is required to merge objects into DAS records based
on their common key (block name in this case) and output creation time 
is a time required to write DAS records to disk. Please note that last two
components, aggregation and output creation time, were only applied to
final DAS step and not to individual components.

The individual DAS components (DBS and PhEDEx) shown roughly the same performance.
For example, in case of PhEDEx the total time spent in DAS was 177 seconds, where
majority of the time was spent on retrieval, 83 seconds, and parsing, 29 seconds.
The remaining 65 seconds was spent on remapping, cache insertion, indexing. The
numbers for DBS system were a like. As you can see from Table \ref{DAS_benchmark},
the time spent to read DAS records (last column) from cache was quite
reasonable, roughly 2.5 seconds from both systems. While final time to
get DAS records on disk was below 42 seconds. This was mostly due to I/O operations.

@@

\section{Summary}
We presented new data aggregation service (DAS) developed for High-Energy experiment
at LHC, CERN, Geneva, Switzerland. It was designed to providing caching and
aggregation layer on top of existing relational and non-relation data-service
in mostly real time fashion. All the data were retrieved on demand basis,
while a pre-fetch of most common queries is in our plans. The DAS system reach
first is under commissioning phase right now and prelimiary performance studies
shown it can accommodate needs of CMS experiment. The CMS experiment started to
collect data from November 2009 and we expect to write about PB of data each
year to tape. In addition the Monte-Carlo samples will be produced at this scale.
Based on test studies performed in CMS, it can translate to the order of
$\sim500$GB of meta-data each year. The DAS should sustain such load and provide
interface as data discovery service in CMS.

@@

\bibitem{LumiDB}
https://twiki.cern.ch/twiki/bin/view/CMS/CMS-DMWM-DBS-Luminosity

\bibitem{DQ}
Data-Quality DB reference

\bibitem{Overview}
https://cmsweb.cern.ch/overview/

@@

>>>>

The DAS core communicates with different data-services and retrieve
data on demand. This was done by invoking appropriate data-service APIs
upon user requests. All user queries were written to Analytics DB
for further analysis. The data-service output where re-mapped into
DAS notations and if necessary aggregated inside of DAS cache.
Since we didn't know a-priory what kind of data users will ask, 
what kind of queries they will place against the system and what
level of aggregation will be applied the choice
of RDMS as DAS back-end was problematic\footnote{In addition DAS 
didn't impose any requirements for transactions and persistency of the data.} 
and we look around for other options. Through the analysis of available IT solutions, 
such as file-based and memory caches, 
key-value databases, documented-oriented databases we made our choice in favor 
of the last technology. Among them we evaluated CouchDB \cite{CouchDB} and 
MongoDB \cite{MongoDB}. Both of them provide schema free document
storage, replication and fail-over features, but our choice in favor of 
MongoDB was obvious because its support of dynamic queries, 
full indexies, including on inner objects and embedded arrays,
and auto-sharding. Our prelimiary benchmarks shown that it can sustain
the desired load and size for our meta-data information. We use MongoDB as a back-end 
for the three main components: Mapping, Analytics and DAS cache databases.

DAS Mapping DB holds mapping between data-service APIs and DAS keys. 
Upon identification of APIs which will participate in DAS activity 
we store to Mapping DB the API name, input parameters, their type and accept 
values as regular expression patterns as well as mapping between data-service 
notations into DAS keys. This allows to map user input query parameters into
data-service API metrics as well as map back data-service output into DAS names.
For instance, the DBS system \cite{DBS}
used {\it logical\_file\_name} notation, while in PhEDEx
\cite{PhEDEx}
the same entity was named as {\it lfn}. Since both of them were referred to
logical file name, we map both notations into DAS key {\it file}.\footnote{We
also deal with physical file names, but still used {\it file} to refer to them in
DAS notation, since simple regular expressions were used to identify {\it file}
meaning from the provided value.}
For each service we provide a mapping from its own data-format (JSON, XML, etc.) into
common data-format used in DAS.\footnote{We used JSON as a DAS data format.}
We also store a mapping between DAS keys and data-service API names to
establish appropriate calls to data-service upon provided DAS query. For example,
when user specifies {\it file} in their DAS query, we invoke DBS listFiles and PhEDEx
fileReplicas APIs.

The DAS Analytics DB plays special role. It collects information
about user requests placed to the system. Each request was recorded. Upon its
decomposition into set of selection keys and conditions we also record
API name and API input parameters and keep updating their counter for repeated
calls. This allows to keep track of frequency of API calls and establish pre-fetch
strategy for most requested APIs.

The DAS caching system was used to store data-service results and aggregated
information in form of DAS records. As it shown in figure \ref{DAS_cache} 
the DAS cache server updates Analytics DB, requests and retrieve data from DAS cache
and aggregate data over there. The independent set of daemons, DAS robots, are
constantly monitor Analytics DB for most popular requests and pre-fetch most
common data in DAS cache.

@@

Apart from previous DBS-QL syntax we decided to fully support free text-based queries for end-users.
All of them were transformed into MongoDB query syntax, who was used internally in
DAS core layer. It represents a set of selection keys and conditions as a JSON document.
For example, a query
{\it dataset site=T1\_CH\_CERN}
where transformed into 
{\it \{"fields" : ["dataset"], "spec" : \{"site" : "T1\_CH\_CERN"\}\}}.
Such syntax was perfect fit for us, since it allows us to store all queries
as JSON documents into Analytics DB for further analysis. It is worthwhile to note that
MongoDB query syntax is very reach. It supports advanced queries with 
boolean expressions, conditional operators, regular expression, etc. providing
a real power for our end-users.

\subsection{Data-services and aggregation}
DAS provided a pluggable interface to add new CMS data-services. 
Each data-service plugin consists of its DAS map and individual data parser. 
The data-service DAS map consists of information about data-service APIs,
input and output parameters, data-service
format and their mapping to DAS keys as well as relation to other sub-systems. 
The information was stored into DAS Mapping DB. Below you can see a typical
record of SiteDB \cite{SiteDB} data-service:
\begin{verbatim}
{"api": {"params": {"name": ""}, "name": "CMSNametoSE"}, 
 "api2das": [{"pattern": "re.compile('^T[0-3]_')", 
              "api_param": "name", 
              "das_key": "site"}], 
 "_id": "4ae1f8dfe2194e4581000014", 
 "system": "sitedb", 
 "daskeys": [{"pattern": "re.compile('^T[0-3]_')", 
              "key": "site", 
              "map": "site.name"}]}
\end{verbatim}
It contains information about API name and parameters, mapping between API input
parameter and DAS key, CMS system name and set of DAS keys this API is covering.
We also used regular expression patterns, evaluated by python during run time
execution, to verify provided set of input paremters and DAS keys.

The DAS Mapping DB becomes an authorative
source of information about all data-services participating in DAS and their
relationships. Based on this information we were able to retrieve appropriate data
from different data-services, re-map them into DAS notations and
aggregate them on demand.
For instance, the Data Bookkeeping System (DBS) and data location system (PhEDEx)
provided information about CMS data files. In former case, DBS holds information about
files and their relationship to other physics objects, such as run and luminosity, 
while PhEDEx stores information about the file location, such as sites which
holds copy of the files and details about storage elements.
Upon a user request to find a file, we were able to identify appropriate set of
DBS and PhEDEx APIs, look-up data, re-format their output to DAS notations and
store results into DAS cache.
Once data were retreived from data-services an additional step to aggregate
matched objects in DAS cache were applied. For instance, both data-service outputs 
contained a file name in them. We use it as a common key to merge records in DAS cache.
This merged (aggregated) results represented DAS data objects. Each DAS data
object contains a standard DAS header, which identify look-up time,
request url, input parameters and API method as well as response version, expiration
time and checksum of the output. Here is an example of individual DAS record:
\begin{verbatim}
{"_id": "4ae1e7ffe2194e4243000003", 
 "site":[{"samname": "CERN-PROD", "name": "T1_CH_CERN"}, 
         {"sitename": "CERN", "name": "T1_CH_CERN"}], 
 "das": [{"qhash": ["ed2b73067794d443ed56756d9b94f7bc"], 
          "ctime": 0.95439004898071289, "expire": 1256362175.7692671, 
          "url": "https://cmsweb.cern.ch/sitedb/json/index/CMStoSAMName", 
          "timestamp": 1256318975.7692671, "system": ["sitedb"], 
          "api": ["CMStoSAMName"], "version": "", "selection_keys": []}, 
         {"qhash": ["ed2b73067794d443ed56756d9b94f7bc"], 
          "ctime": 1.2158100605010986, 
          "api": ["CMStoSiteName"], 
          "url": "https://cmsweb.cern.ch/sitedb/json/index/CMStoSiteName", 
          "timestamp": 1256318974.7774379, "system": ["sitedb"], 
          "expire": 1256362174.7774379, "version": "", "selection_keys": []}
        ]
}
\end{verbatim}
It represents aggregated DAS record between two SiteDB APIs, CMStoSAMName and
CMStoSiteName, where API results are merged under {\it site} key. The information
under {\it das} key shows DAS header, which contains sufficient amount of
information about API name, location, execution time and lifetime of the data.
Using this information was valuable contribution in debugging process of
data-services themselves, e.g. identification of not-matched records, 
latency studies, etc.

We used REST model \cite{REST} for all data requests to DAS.
It allows us to share core library and other components during development of 
various DAS clients, e.g. CLI, web interface, robots, etc. and hides
complexity of access to individual data-services and security policies.
For instance, to access on-line Run Summary information we were forced to use
Grid certificates. Instead of propagating user certificate to data-service
the DAS works as a proxy and used host certficiate to serve such requests.

\section{Results\label{Results}}
DAS was deployed to work with the following CMS data-services:
\begin{itemize}
\item DBS \cite{DBS}, Data Bookkeeping System which collect information
about CMS meta-data;
\item PhEDEx \cite{PhEDEx}, Physics Experiment Data Export project which
provides the data placement and the file transfer system for the CMS experiment;
\item SiteDB \cite{SiteDB} is a catalogue of all CMS computing Tiers. 
It records the CMS resources at the site, the resources pledged for the 
future, and it keeps track of CMS personnel at each site, including the 
roles and duties they fulfill in the collaboration;
\item RunSummary DB \cite{RunSummary} collects information about run and triggers
conditions during data-taking;
%\item LumiDB \cite{LumiDB} collects information about Luminosity conditions;
%\item DQ \cite{DQ}, Data Quality data-service which collects information
%about detector conditions during data taking;
%\item Overview \cite{Overview}, collects information about CMS
%transfer rate, etc.;
\item Dashboard \cite{Dashboard} project for LHC experiments aims to 
provide a single entry point to the monitoring data collected from the 
distributed computing systems of the LHC virtual organization.
\end{itemize}
Each data-service has its own scope, size and lifetime. For instance the data
in PhEDEx are transient due to constant migration of CMS data with daily DB size of
around few tens of GBs. The DBS system were divided into dozen of individual instances, 
who has been used by different production teams for various bookkeeping tasks.
Figure \ref{db_size} shows current size of meta-data from aforementioned data-service.
\begin{figure}[htb]
\centering
\includegraphics[width=100mm]{db_size.pdf}
\caption{
Database size of CMS data-services participating in DAS in November 2009, before
data taking.
}
\label{db_size}
\end{figure}
%Typical
%size of individual DBS instance was around of fwe GBs, with overall DBS size of
%around of 50 GBs.\footnote{This include table and index sizes.}

At the moment we don't have exact numbers for amount of data which will stored to DAS,
but based on test performed during CRAFT09 \cite{CRAFT09}
exercise we produced 0.3PB of data which
translate into $\sim50$ GB of meta-data stored into DBS system. Therefore
we can anticipate to collect $\sim500$GB of meta-data per year during data-taking period
at LHC.
%In CRAFT09 acquisition era, Tier-0 produced a total of 6.03
%billion events, 330TB in 32 datasets:
%AlCa/Calibration RAW: 926 million events, 29TB
%Physics RAW: 1.24 billion events, 147TB
%PromptReco: 1.24 billion events, 111TB
%Bulk ALCARECO: 2.34 billion events, 11TB
%Express FEVT: 105 million events, 24TB
%Express ALCARECO: 153 million events, 0.99TB
%HLTMON FEVTHLTALL: 31 million events, 7.4TB
%Above event numbers have signicant overlap by design, 
%2.2 billion unique events, including 524 million cosmic muon
%triggers
% 0.3PB - 50GB meta-data in DBS
% 3PB/y - x GB

@@

We measured elapsed time as the following:
\begin{itemize}
\item[]
{\it
Elapsed time = retrieval time + parsing time + remapping time 
        + cache insertion/indexing time 
        + (aggregation time) + (output creation time)
}
\end{itemize}
Here the {\it retrieval time} is a time required to access data from remote data-services,
{\it parsing time} is a time required to read and parse received data, {\it remapping time}
is a time to convert notations used by data-service to DAS ones for every object
we parse, {\it cache insertion} and {\it indexing time} represents time spend to add objects into
the DAS cache, {\it aggregation time} is required to merge objects into DAS records based
on their common key (block name in this case) and {\it output creation time}
is a time required to write DAS records to disk. Please note that last two
components, {\it aggregation} and {\it output creation time}, were only applied to
final DAS step and not to individual DAS components.

The individual DAS components (DBS and PhEDEx) shown roughly the same performance.
For example, in case of PhEDEx the total time spent in DAS was 177 seconds, where
majority of the time was spent on retrieval, 83 seconds, and parsing, 29 seconds.
The remaining 65 seconds was distributed among remapping, cache insertion, indexing. The
numbers for DBS system were alike. As you can see from Table \ref{DAS_benchmark},
the time spent to read DAS records (last column) from the cache was quite
reasonable, roughly 2.5 seconds from both systems. While final time to
get DAS records on disk was below 42 seconds. This was mostly due to I/O operations.

@@

\section{Summary}
We presented new data aggregation service (DAS) developed for CMS High-Energy experiment
at LHC, CERN, Geneva, Switzerland. It was designed to provide caching and
aggregation layer on top of existing relational and non-relation data-service
mostly in real time fashion. All the data were retrieved on demand basis,
while data pre-fetching of most common queries is in our plans. The DAS system 
is under commissioning phase right now and prelimiary performance studies
show it can accommodate needs of CMS experiment as general purpose meta-data
service. The CMS experiment is started to
collect data from November 2009 and we expect to write about PB of data each
year to tape. In addition, the Monte-Carlo samples will be produced at this scale.
Based on test studies performed in CMS, it can translate to the order of
$\sim500$GB of meta-data each year. The DAS should sustain such load and provide
interface as data discovery service in CMS.

@@

%\bibitem{LumiDB}
%https://twiki.cern.ch/twiki/bin/view/CMS/CMS-DMWM-DBS-Luminosity

%\bibitem{DQ}
%Data-Quality DB reference

%\bibitem{Overview}
%https://cmsweb.cern.ch/overview/

@@

\bibitem{CRAFT09}
https://twiki.cern.ch/twiki/bin/view/CMS/CRAFT09AnalysisInfo
